---
title: "Multivariate analysis of the exposome with mixOmics"
subtitle: "The HELIX case study"
author: "Kim-Anh Lê Cao"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_float: yes
    toc_depth: '3'
  html_notebook:
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: 3
editor_options:
  markdown:
    wrap: 72
  chunk_output_type: console
bibliography: ["bibliography.bib"]
biblio-style: apalike
link-citations: true
---

```{=html}
<!--
Show / hide answers to exercises.
Code adapted from: https://chrisbeeley.net/?p=1104
-->
```
```{=html}
<script>
function myFunction(id) {
    var x = document.getElementById(id);
    if (x.style.display === "none") {
        x.style.display = "block";
    } else {
        x.style.display = "none";
    }
}
</script>
```
```{=html}
<style>
div .info {
  margin: auto;
  background-color: #EAF0FB;
  width: 95%;
  padding: 10px;
}

/* The citation ISGlobal requires verbatim. Bootstrap styles a blockquote at
   17.5px, which is larger than body text and reads oddly for what is really a
   block of boilerplate, so it is set smaller and tighter here. The
   word-breaking keeps the long Figshare and GitHub URLs inside the box. */
div .required-citation {
  margin: 1.2em auto;
  width: 95%;
  padding: 12px 18px;
  background-color: #F7F7F7;
  border-left: 4px solid #585858;
  font-size: 0.85em;
  line-height: 1.55;
  color: #333333;
  overflow-wrap: break-word;
  word-break: break-word;
}
div .required-citation p {
  margin: 0;
}
</style>
```
```{r setup, include=FALSE}
# Smaller images for pdf
# knitr::opts_chunk$set(out.width="50%")
options(width=80)
```


```{r 01-options, include=FALSE}
library(knitr)

# Show or hide the output of every code chunk.
#
#   TRUE  - print the results and draw the figures. Use this to check the
#           material renders and reads correctly before a workshop.
#   FALSE - show the code and nothing else. This is the version participants
#           work from: they run each line themselves and read the results in
#           their own console.
#
# Inline `r ...` results in the exercise answers are unaffected and always
# appear, whichever way this is set.
show.results <- FALSE

# global options
knitr::opts_chunk$set(dpi = 100, echo = TRUE, warning = FALSE, message = FALSE, eval = TRUE,
                      results  = if (show.results) 'markup' else 'hide',
                      fig.show = if (show.results) 'asis'   else 'hide',
                      fig.width = 9, fig.height = 7.5, fig.align = 'center', out.width = '70%')

colorize <- function(color, x) {
  if (knitr::is_html_output()) {
    htmlcolor = "black"
    if(color == "blue"){
      htmlcolor = "#388ECC"
    }
    if(color == "orange"){
      htmlcolor = "#F68B33"
    }
    if(color == "grey"){
      htmlcolor = "#585858"
    }
    if(color == "green"){
      htmlcolor = "#009E73"
    }
    if(color == "pink"){
      htmlcolor = "#CC79A7"
    }
    if(color == "yellow"){
      htmlcolor = "#999900"
    }
    if(color == "darkred"){
      htmlcolor = "#CC0000"
    }
    sprintf("<span style='color: %s;'>%s</span>", htmlcolor, x)
  } else {
    # Any non-HTML output, such as the pdf_document format advertised in the
    # YAML above. Return the text unchanged: without this branch the function
    # returns NULL invisibly and every coloured word vanishes from the
    # sentence that contains it.
    x
  }
}
```
# Let's get started {#00}

## Installation {#00:install}

First, install the latest version of `mixOmics` from Bioconductor [@Roh17]:

```{r 00-install-bioc, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
 BiocManager::install("mixOmics")
```


`mixOmics` will also install the packages it imports:
`igraph`, `rgl`, `ellipse`, `corpcor`, `RColorBrewer`, `plyr`,
`parallel`, `dplyr`, `tidyr`, `reshape2`, `methods`, `matrixStats`,
`rARPACK` and `gridExtra`.

**For Mac users:** if the imported package `rgl` does not install, install
[XQuartz](https://www.xquartz.org) first and then try again.

## Load the package {#00:load-package}

```{r 00-load, message=FALSE}
library(mixOmics)
```

Check that the package loads without an error. If the error mentions `rgl` on
a Mac, see the XQuartz note above.

## The HELIX data {#00:helix}

HELIX (Human Early-Life Exposome) measured four molecular layers and more than
200 environmental exposures in the same children. The children came from six
population-based birth cohorts in France, Greece, Lithuania, Norway, Spain and
the United Kingdom [@Vri14; @Mai18]. Mother–child pairs were followed from
pregnancy, and the children were re-examined at age 6 to 11.

This practical uses a processed and normalised subset of the data released by
ISGlobal for its Exposome Data Challenge. The data were processed and stored in a single list,
`helix`, containing 896 children.
The samples from the same children are in the **same order**, which is required to perform data integration.

| element | what it contains |
|---|---|
| `helix$omics` | four matrices, 896 rows each: `proteome` (36 circulating proteins), `serum` (177 metabolites), `urine` (44 metabolites) and `transcriptome` (1,000 transcripts) |
| `helix$exposure` | nine families of postnatal environmental exposures, from 4 to 15 variables each, 81 variables in total |
| `helix$outcomes` | characteristics of the children, stored as factors |
| `helix$covariates` | five numeric characteristics of the children |
| `helix$annotation` | feature annotation for each omics block, and the exposure codebook |
| `helix$provenance` | where the data came from, and processing steps applied |

The [repository README](https://github.com/mixOmics-org/mxoorg-workshop-exposome-helix#the-data)
describes how the extract was prepared and what processing was applied to each
omics block.

## Load the data

The dataset is available in the repository's `Data/` folder.
First, save the .RData into your working directory folder.
Second, set the working
directory to the folder containing this file. In RStudio, open this `.Rmd` and
select **Session > Set Working Directory > To Source File Location**.
Then load
the data. The command may look like this, depending on where the data are stored:

```{r 00-read-data, eval = FALSE}
load("helix.RData")
```

```{r 00-read-data-render, include = FALSE}
# Not shown to participants. The rendered document is built from the repository
# layout, where Data/ sits beside practical/, so it loads the file from there.
# Participants who have saved helix.RData into their own working directory use
# the line above.
load(file.path("..", "Data", "helix.RData"))
```

Before running an analysis, inspect the objects you have loaded:

```{r 00-inspect}
names(helix)

sapply(helix$omics, dim)     # four molecular blocks, samples in rows
sapply(helix$exposure, ncol) # nine families of environmental exposures

summary(helix$outcomes)      # characteristics of the children, all factors
str(helix$covariates)        # and the numeric ones
```

The exposure variables have abbreviated names.
The codebook gives a description of each variable:

```{r 00-codebook}
head(helix$annotation$codebook)

# what is hs_sumPCBs5_cadj_Log2?
helix$annotation$codebook["hs_sumPCBs5_cadj_Log2", ]
```

## Quick start in `mixOmics` {#00:quick-start}

In mixOmics, each analysis follows the same workflow:

1. Run the method.
2. Plot the samples.
3. Plot the variables.

Principal Components Analysis (PCA) summarises the main patterns of variation in
a data set using a small number of components.
To try this workflow, first store
the proteome in its own object:

```{r 00-quick-load}
proteome <- helix$omics$proteome   # the data set mixOmics calls X
dim(proteome)                      # 896 children, 36 proteins
```

Now run the three steps:

```{r 00-quick-pca}
MyResult.pca <- pca(proteome) # 1 Run the method PCA
plotIndiv(MyResult.pca)       # 2 Plot the samples from a PCA result
plotVar(MyResult.pca)         # 3 Plot the variables from a PCA result
```

For now, hold off on interpreting the plots, which we will do in the next sections.
The same three steps recur throughout the practical as the methods become more complex.

# 1. PCA on the HELIX case study {#01}

We consider the following two data sets:

-  the **proteome** contains 36 circulating proteins, largely inflammatory markers and cytokines.

-  the **organochlorines** data set contains nine persistent chemical pollutants
measured in each child's serum. The
[repository README](https://github.com/mixOmics-org/mxoorg-workshop-exposome-helix#the-data)
describes how the datasets were prepared.

## Load the data

We store each omics dataset as an object. In the sections that follow, `proteome` and
`organochlorines` take the roles of $\boldsymbol X$ and $\boldsymbol Y$,
respectively. This is the notation used in the `mixOmics` arguments and
documentation.

```{r 01-load, message=FALSE}
proteome        <- helix$omics$proteome            # 36 circulating proteins
organochlorines <- helix$exposure$organochlorines  # 9 persistent pollutants

dim(proteome)   # check dimension
dim(organochlorines)
```

## Exploration of the proteomics dataset with PCA {#01:pca-proteome}

PCA is an unsupervised method: the aim is to identify the main sources of variation
without including any information about the children in the analysis. We can then colour the sample
plot by a known characteristic, such as sex or recruitment cohort, to see
whether that characteristic helps explain the pattern. Here, we centre the data
(the default) and scale each protein before fitting the PCA:

```{r 01-pca-proteome}
pca.prot <- pca(proteome, ncomp = 3, scale = TRUE)

pca.prot

plot(pca.prot)   # screeplot

# simple version
plotIndiv(pca.prot,
          pch = 1,    # pch is to show symbols
          title = 'HELIX proteome, PCA comp 1 - 2')

# advanced version
plotIndiv(pca.prot,
          group = helix$outcomes$sex, # asking to color according to sex
          ind.names = FALSE,          # not showing the sample names
          legend = TRUE, legend.title = 'Sex',
          title = 'HELIX proteome, by sex')
```

Recolour the same sample plot twice more, first by recruitment cohort and then
by BMI category. Only the `group` argument changes:

```{r 01-pca-proteome-recolour}
plotIndiv(pca.prot, group = helix$outcomes$cohort, ind.names = FALSE,
          legend = TRUE, legend.title = 'Cohort', ellipse = TRUE,
          title = 'HELIX proteome, by recruitment cohort')

plotIndiv(pca.prot, group = helix$outcomes$bmi_cat, ind.names = FALSE,
          legend = TRUE, legend.title = 'BMI category', ellipse = TRUE,
          title = 'HELIX proteome, by BMI category')
```

## Exploration of the exposure dataset organochlorines with PCA {#01:pca-oc}

We apply the same analysis to the nine organochlorines. The function and arguments
are unchanged:

```{r 01-pca-oc}
pca.oc <- pca(organochlorines, ncomp = 3, scale = TRUE)

pca.oc

plot(pca.oc)

plotIndiv(pca.oc, group = helix$outcomes$cohort, ind.names = FALSE,
          legend = TRUE, legend.title = 'Cohort', ellipse = TRUE,
          title = 'Organochlorines, by recruitment cohort')

plotIndiv(pca.oc, group = helix$outcomes$bmi_cat, ind.names = FALSE,
          legend = TRUE, legend.title = 'BMI category', ellipse = TRUE,
          title = 'Organochlorines, by BMI category')
```

## Exercise 1: Interpretation

Run the code above, line by line.

1. How many components would you retain in each PCA, and why?

2. Compare the proteome plots coloured by sex, cohort and BMI category. Which
   characteristic explains most of the variation? Does the same characteristic
   come first in the organochlorines?

3. The same characteristic structures the first component of both data sets.
   What does this imply when relating the proteome to the organochlorines? What
   would you check before interpreting an association between them?

4. *A more advanced question.* A sample plot shows whether a characteristic
   separates the children. To quantify this pattern, we can regress a component on a
   characteristic and examine R², the proportion of variation in that component
   explained by the characteristic:

   ```r
   # proteome: component 1, then component 2
   summary(lm(pca.prot$variates$X[, 1] ~ helix$outcomes$cohort))$r.squared
   summary(lm(pca.prot$variates$X[, 2] ~ helix$outcomes$bmi_cat))$r.squared

   # organochlorines: component 1, then component 2
   summary(lm(pca.oc$variates$X[, 1] ~ helix$outcomes$cohort))$r.squared
   summary(lm(pca.oc$variates$X[, 2] ~ helix$outcomes$cohort))$r.squared
   ```

   Do the four values agree with the ranking suggested by the sample plots?

<button onclick="myFunction(&#39;q1&#39;)">

Show solutions

</button>

::: {#q1 style="display:none"}

### Answers

1. For the proteome, the first two components explain `r round(pca.prot$cum.var[2]*100, 1)`%
   of the total variance, while the third adds only
   `r round(pca.prot$prop_expl_var$X[3]*100, 1)`%. Using the elbow method, we
   would retain 2 components. For the organochlorines, the first component alone
   explains `r round(pca.oc$prop_expl_var$X[1]*100, 1)`%, and the screeplot
   reaches its elbow immediately afterwards. The amount of explained variance differs because the
   nine organochlorines largely measure one underlying pattern: the congeners
   co-accumulate in the same children. The 36 proteins do not.

2. Neither data set separates by `r colorize("blue", "sex")`; the two groups
   overlap in both plots. **Cohort** produces the clearest separation in both
   data sets and is more visible in the proteome's first component than in the
   organochlorines'. The second components differ. In the proteome, component 2
   reflects BMI category; in the organochlorines, it still reflects cohort. The
   `r colorize("grey", "thinness")` category contains only 9 of the 896
   children and is therefore almost invisible. Check group sizes with
   `summary(helix$outcomes)` before interpreting colour patterns.

3. In both data sets, the largest single source of variation is **where the child
   was recruited** (cohort), rather than a characteristic of the individual child. For
   the proteomics data, this may reflect collection and handling as well as
   biology. A model could therefore associate the proteome with organochlorines
   by capturing differences between cohorts rather than differences between
   children. To check this, we could stratify by cohort, then ask whether the association remains within cohorts.
   Code for this check appears in *To go further*.

4. The values agree with the plots. Cohort explains R² =
   `r round(summary(lm(pca.prot$variates$X[,1] ~ helix$outcomes$cohort))$r.squared, 2)`
   of proteome component 1 and
   `r round(summary(lm(pca.oc$variates$X[,1] ~ helix$outcomes$cohort))$r.squared, 2)`
   of organochlorine component 1. The second components differ: BMI category
   explains
   `r round(summary(lm(pca.prot$variates$X[,2] ~ helix$outcomes$bmi_cat))$r.squared, 2)`
   of proteome component 2, while cohort explains
   `r round(summary(lm(pca.oc$variates$X[,2] ~ helix$outcomes$cohort))$r.squared, 2)`
   of organochlorine component 2. These R² values are descriptive. They state
   how much of a component a characteristic explains, not whether that
   component is important.
:::

<!-- end solutions -->

# 2. PLS1 — regressing one exposure on one omics dataset {#02}

We now ask whether variation in the proteome can explain variation in a **single
exposure**. Partial Least Squares regression with one response is called PLS1
[@Lec08].

## The response

The response is `hs_sumPCBs5_cadj_Log2`: the summed serum concentration of five
polychlorinated biphenyls (PCBs), adjusted for lipids and log2 transformed.
PCBs are **persistent organic pollutants**. Although these
industrial compounds were banned decades ago, they resist metabolism and are
measurable in every child in this cohort. They are also strongly **lipophilic**:
they accumulate in body fat and circulate bound to serum lipids, which is why
their concentration is expressed per unit of lipid. Their documented effects
include immune and endocrine toxicity, while the proteome used here consists
largely of circulating inflammatory markers. This makes summed PCBs a relevant
exposure to examine against these proteins.

We use the sum of five congeners rather than a single congener. Section 1 showed
that the organochlorines are largely one-dimensional because they co-accumulate.
The sum represents this shared pattern and is less noisy than any one congener.

## Set the response variable

We select the appropriate column of the
organochlorine block as the response:

```{r 02-set-y, message=FALSE}
Y <- organochlorines[, "hs_sumPCBs5_cadj_Log2"]

length(Y)
summary(Y)

hist(Y, breaks = 30, xlab = "summed PCBs (log2, lipid-adjusted)",
     main = "The response")
```

The response is a **continuous** measurement rather than a group label.
This distinguishes regression from discriminant (classification) analysis.

## First pass: PLS1

Fit a PLS model with three components in regression mode:

```{r 02-pls}
pls.prot <- pls(proteome, Y, ncomp = 3, mode = "regression")

pls.prot
```

The model contains three components, but they may not all improve prediction.
Use cross-validation with `perf()` and the Q2 statistic to decide how many to
retain [@Sta87]:

```{r 02-perf}
perf.pls.prot <- perf(pls.prot, validation = "Mfold", folds = 5,
                      nrepeat = 10,
                      seed = 42)   # for a reproducible cross-validation

plot(perf.pls.prot, criterion = "Q2")

perf.pls.prot$measures$Q2$summary
```

<p>
::: info
**ⓘ We use Q2 to choose the number of components**

Q2 compares the model's predictions for held-out children with predictions based
only on the average. We retain a component if its Q2 is above the horizontal line
at 0.0975, and exclude any component below the line. Q2 indicates **how
many components to retain**. It does not measure the strength or importance of
the relationship.

For reproducible cross-validation, pass the seed through the `seed` argument of
`perf()`, as shown above.
:::
</p>

The result supports **one component**. Component 1 has a Q2 of
`r round(perf.pls.prot$measures$Q2$summary$mean[1], 3)`, above the line.
Component 2 has a Q2 of
`r round(perf.pls.prot$measures$Q2$summary$mean[2], 3)`, below the line, and
component 3 is negative. A negative value means that the component predicts
held-out children worse than the training average. One component is the usual
result in PLS1, where we have a single response. However, we retain a second
component for plotting, and will primarily focus our interpretation on the first
component.

## Examine the model

First, plot the children in the space spanned by the proteome components and
colour them by cohort. We then inspect the protein loadings for component 1.

```{r 02-pls-plots}
plotIndiv(pls.prot, rep.space = 'X-variate',
          group = helix$outcomes$cohort, ind.names = FALSE,
          legend = TRUE, legend.title = 'Cohort',
          title = 'PLS1 on the HELIX proteome, comp 1 - 2')

# block = 'X' plots the proteome loadings only. The Y block holds the single
# response column, so plotting it would draw a chart with one bar.
plotLoadings(pls.prot, comp = 1, ndisplay = 20, block = 'X',
             title = 'Proteome loadings, component 1')
```

The loading plot shows which proteins contribute to component 1, but not the
direction of their relationship with the exposure. To do this, we plot the **observed** PCB
concentration against the component, add a fitted line and calculate the correlation between the component and the response:

```{r 02-scatter}
t1 <- pls.prot$variates$X[, 1]   # component 1 of the proteome

plot(t1, Y,
     pch = 16, col = "#388ECC66",
     xlab = "proteome, component 1",
     ylab = "summed PCBs (log2, lipid-adjusted)",
     main = "Observed exposure against the proteome component")
abline(lm(Y ~ t1), lwd = 2)

cor(t1, Y)
```

The vertical axis shows the measured PCB concentration in its original analysis
units. The correlation between the component and the exposure is
`r round(cor(pls.prot$variates$X[,1], Y), 2)`.

## Interpretation

Section 1 showed that cohort structures both blocks. Some of the horizontal
spread therefore represents differences *between* cohorts rather than between
children within a cohort, and the fitted line passes partly through the six
cohort means. The association is partly between cohorts. *To go further* shows
how to refit the model after removing cohort from the predictor and response.

## Sparse PLS1 for variable selection

Sparse PLS fits the components and selects the set of proteins that contribute to them
within the same analysis.

<p>
::: info
**ⓘ Sparse means zeroes in the loading vectors**

In ordinary PLS, every protein can contribute to a component, as all loading
weights are non-zero. Sparse PLS uses
lasso penalisation to set some loading values to zero [@Lec08; @Lec11]. Only
proteins with non-zero loadings contribute to the component, so the method also
performs variable selection.
:::
</p>

We need to specify how many proteins to select on each component:

```{r 02-spls}
spls.prot <- spls(proteome, Y, ncomp = 2, mode = "regression",
                  keepX = c(10, 10) # select 10 proteins on each of the 2 components
                  )

plotIndiv(spls.prot, rep.space = 'X-variate',
          group = helix$outcomes$cohort, ind.names = FALSE,
          legend = TRUE, legend.title = 'Cohort',
          title = 'sparse PLS1 on the HELIX proteome, comp 1 - 2')

plotLoadings(spls.prot, comp = 1, block = 'X',
             title = 'Selected proteins, component 1')

plotVar(spls.prot, cutoff = 0.3)  # variable plot (correlation circle plot)

selectVar(spls.prot, comp = 1) #extracts the list of variables selected, and their loading weights

# the same scatter plot, on the sparse model
t1.sparse <- spls.prot$variates$X[, 1]
plot(t1.sparse, Y, pch = 16, col = "#F68B3366",
     xlab = "proteome, component 1 (10 proteins)",
     ylab = "summed PCBs (log2, lipid-adjusted)")
abline(lm(Y ~ t1.sparse), lwd = 2)

cor(t1.sparse, Y)
```

## Exercise 2: Interpretation

Run the code above, line by line.

1. Which proteins have the strongest loadings on component 1, and do their
   loadings have the same sign? Use the scatter plot to determine whether a
   child with a high PCB burden has higher or lower levels of these proteins.
   Is this biologically plausible?

2. What changes when you move from PLS1 to sparse PLS1? Compare the loading
   plot, correlation circle plot and component–exposure correlation. Then change
   `keepX = c(10, 10)` to `c(5, 5)` and `c(30, 30)` and compare the results.

<button onclick="myFunction(&#39;q2&#39;)">

Show solutions

</button>

::: {#q2 style="display:none"}

### Answers

1. The largest loadings are `r colorize("blue", "PAI1")`, followed by
   `r colorize("blue", "EGF")`, `r colorize("blue", "IL4")`,
   `r colorize("blue", "IL2")` and `r colorize("blue", "MIP1alfa")`. They
   **all have the same sign**. Component 1 is therefore a weighted average of a
   coherent group of cytokines and growth factors, rather than a contrast
   between two groups of proteins. The scatter plot establishes the direction;
   `cor(proteome[, "PAI1"], Y)` confirms it with a value of
   `r round(cor(proteome[, "PAI1"], Y), 2)`. Children with a **higher** PCB
   burden have **lower** circulating levels of these proteins. This pattern is
   consistent with reported immunosuppressive effects of these compounds, but
   it remains an association rather than a mechanism. The cohort caveat from
   Section 1 also remains.

2. The loading plot now contains only the 10 selected proteins rather than all
   36 proteins ranked by contribution. In the correlation circle plot, the
   selected proteins are closer to the circle and better aligned with the axes.
   The component–exposure correlation changes little:
   `r round(cor(spls.prot$variates$X[,1], Y), 2)` with 10 proteins, compared
   with `r round(cor(pls.prot$variates$X[,1], Y), 2)` with all 36. Selecting
   10 proteins therefore retains almost all of this association.
   Changing `keepX` to 5 or 30 alters the loading plot more than the correlation.
   The selected proteins remain in a similar order, and the sparse model
   approaches the full PLS model as `keepX` approaches 36. In a two-block model,
   `tune.spls()` can select `keepX` by cross-validation. Tuning is outside the
   scope of this workshop; see the
   [sPLS documentation](http://mixomics.org/methods/spls/).
:::

<!-- end solutions -->

# 3. Block sPLS to integrate several omics data sets

Section 2 related one omics dataset to one response. HELIX measured four omics
blocks in the same children. `block.spls` can relate all four blocks to the same
response in a single model [@Ten14]. It is the regression counterpart of
DIABLO [@Sin19], with a continuous response block in place of a class label.

The organochlorines will be the response *matrix*, and the other omics datasets
(proteome, serum metabolome, urine metabolome and transcriptome) the predictors.

<p>
::: info
**ⓘ Align the samples across all omics blocks**

Every block must contain the same individuals in the same order. Each component
has one value for each of the $N$ individuals, and our method measures covariance
between these aligned values. If the rows refer to different individuals, the
estimated relationships between blocks are meaningless.
:::
</p>

For the discriminant version, which relates several omics blocks to a
*categorical* outcome, see the DIABLO analysis in the
[companion practical](https://github.com/mixOmics-org/mxoorg-workshop-intro-mixomics).

## Set up the omics blocks

We store the data matrices in a list, $\boldsymbol X$, and name every element. The
names will be matched to the `keepX` parameter during variable selection. The
response belongs in the **same list**; the `indY` argument tells `block.pls()`
which element is the response.

```{r 03-blocks}
X.list <- list(proteome        = helix$omics$proteome,
               serum           = helix$omics$serum,
               urine           = helix$omics$urine,
               transcriptome   = helix$omics$transcriptome,
               organochlorines = helix$exposure$organochlorines)

sapply(X.list, dim)

names(X.list)   # organochlorines is element 5: that is what indY = 5 refers to
```

## Set the design matrix

The design matrix has one row and one column for each block. Each off-diagonal
entry is a value between 0 and 1 that controls the strength of the relationship
the model seeks between a pair of blocks. A value of 1 asks the two blocks to
agree as strongly as possible; a value of 0 only includes a very small level of integration between them during the optimisation procedure.

Here, we link each omics block to the organochlorines with a value of 1, and to
the other omics blocks with a value of 0.5. This focuses the model on covariance
between each predictor block and the response. Linking the predictor blocks to
one another, as the 0.5 entries do, trades some covariance with the response for
agreement among the predictor blocks. That
choice is available in both regression and classification models. See
[N-integration methods](https://mixomics.org/methods/n-integration-methods/) for
more information about choosing a design.

```{r 03-design}
design <- matrix(0.5, #a medium amount of correlation between omics datasets
        nrow = length(X.list), ncol = length(X.list),
        dimnames = list(names(X.list), names(X.list)))
design["organochlorines", ] <- 1
design[, "organochlorines"] <- 1
diag(design) <- 0

design
```

## Block PLS, then block sPLS

First fit a block PLS model without variable selection:

```{r 03-block-pls}
block.pls.helix <- block.pls(X.list, indY = 5, ncomp = 2,
                             mode = "regression", design = design)
```

Now fit the sparse version. As in Section 2, specify how many variables to keep
for each block and component. Note that `perf()` and tuning are not currently available for
`block.pls` or `block.spls`, so the `keepX` values below are fixed manually
rather than selected by cross-validation. This analysis is therefore
exploratory.

```{r 03-block-spls}
list.keepX <- list(proteome      = c(10, 5), #values can change from one component to the next
                   serum         = c(15, 10),
                   urine         = c(10, 10),
                   transcriptome = c(25, 30))

block.spls.helix <- block.spls(X.list, indY = 5, ncomp = 2,
                               mode = "regression", design = design,
                               keepX = list.keepX)

# sample plots: one panel per block
plotIndiv(block.spls.helix, group = helix$outcomes$cohort,
          ind.names = FALSE, legend = TRUE, legend.title = 'Cohort')

# variable plots:
plotVar(block.spls.helix, cutoff = 0.5, legend = TRUE,
        title = 'Correlation circle, all five blocks')

# circosPlot requires a group argument even though this is a regression and
# there are no groups in the model. Cohort is passed to satisfy it; the
# colouring it produces is decoration, not part of the model.
circosPlot(block.spls.helix, group = helix$outcomes$cohort,
           cutoff = 0.55, line = TRUE,
           size.variables = 0.5, size.labels = 1.2)

# variable selection, one block at a time
selectVar(block.spls.helix, comp = 1, block = "proteome")
selectVar(block.spls.helix, comp = 1, block = "urine")

# no block argument: one panel per block, the response included
plotLoadings(block.spls.helix, comp = 1)
```

<p>
::: info
**ⓘ Why some organochlorines sit outside the correlation circle**

A correlation circle assumes that its two axes are uncorrelated. This is true
for the four predictor blocks, whose successive components are uncorrelated by
construction, but not for the response block. In `mode = "regression"`, the
response is deflated on the predictor components rather than on its own. Its two
components are therefore not forced apart and are
`r round(cor(block.spls.helix$variates$organochlorines[, 1], block.spls.helix$variates$organochlorines[, 2]), 2)`
correlated here.

Because the axes are so similar, a variable strongly correlated with one will
also be strongly correlated with the other. Its plotted point can then fall
beyond a radius of 1. Each coordinate remains a valid correlation, so interpret
the organochlorine positions by their direction rather than their distance from
the origin.

The nearly coincident response components indicate that the organochlorines
contain little second-dimensional information. This agrees with Section 2,
where Q2 retained only one PLS1 component.
:::
</p>

The serum metabolites are named `metab_1`, `metab_2` and so on. They have a
chemical class but no compound identifier because none is available in the
release. Their loadings are therefore not currently biologically interpretable. The
proteome, urine and transcriptome have feature-level annotations:

```{r 03-annotation}
head(helix$annotation$proteome)
head(helix$annotation$urine)
head(helix$annotation$serum)        # Class only, no compound names
head(helix$annotation$transcriptome)
```

## Each predictor block against the response

The following plot contains one panel for each predictor block. Each panel shows
the block's first component on the horizontal axis and the organochlorine first
component on the vertical axis. When the points follow a line, the predictor
block and response share a direction of variation.

```{r 03-panels}
ty <- block.spls.helix$variates$organochlorines[, 1]   # the response variate

par(mfrow = c(2, 2))

tx <- block.spls.helix$variates$proteome[, 1]
plot(tx, ty, pch = 16, col = "#388ECC66", main = "proteome",
     xlab = "predictor variate", ylab = "organochlorine variate")
abline(lm(ty ~ tx), lwd = 2)

tx <- block.spls.helix$variates$serum[, 1]
plot(tx, ty, pch = 16, col = "#388ECC66", main = "serum",
     xlab = "predictor variate", ylab = "organochlorine variate")
abline(lm(ty ~ tx), lwd = 2)

tx <- block.spls.helix$variates$urine[, 1]
plot(tx, ty, pch = 16, col = "#388ECC66", main = "urine",
     xlab = "predictor variate", ylab = "organochlorine variate")
abline(lm(ty ~ tx), lwd = 2)

tx <- block.spls.helix$variates$transcriptome[, 1]
plot(tx, ty, pch = 16, col = "#388ECC66", main = "transcriptome",
     xlab = "predictor variate", ylab = "organochlorine variate")
abline(lm(ty ~ tx), lwd = 2)

par(mfrow = c(1, 1))

# and the four correlations, one line each
cor(block.spls.helix$variates$proteome[, 1],      ty)
cor(block.spls.helix$variates$serum[, 1],         ty)
cor(block.spls.helix$variates$urine[, 1],         ty)
cor(block.spls.helix$variates$transcriptome[, 1], ty)
```

## Exercise 3: Interpretation

Run the code above, line by line.

1. Which predictor blocks track the organochlorines most closely in `plotIndiv`
   and the four scatter plots? Can the four correlations be compared directly?
   Consider the number of features in each block.

2. What does the circos plot show that `plotVar` and `selectVar` do not? What do
   the colours around its outside represent?

3. What can you conclude from the
   analysis, and what can you not conclude?

<button onclick="myFunction(&#39;q3&#39;)">

Show solutions

</button>

::: {#q3 style="display:none"}

### Answers

1. The `r colorize("blue", "proteome")` and
   `r colorize("orange", "serum")` metabolome track the response most closely,
   with correlations of
   `r round(cor(block.spls.helix$variates$proteome[,1], block.spls.helix$variates$organochlorines[,1]), 2)`
   and
   `r round(cor(block.spls.helix$variates$serum[,1], block.spls.helix$variates$organochlorines[,1]), 2)`,
   respectively. The `r colorize("grey", "urine")` metabolome follows at
   `r round(cor(block.spls.helix$variates$urine[,1], block.spls.helix$variates$organochlorines[,1]), 2)`,
   and the `r colorize("green", "transcriptome")` is last at
   `r round(cor(block.spls.helix$variates$transcriptome[,1], block.spls.helix$variates$organochlorines[,1]), 2)`.
   These values should **not** be compared directly. The blocks contain 36, 177,
   44 and 1,000 features, and correlations from a fitted model are not adjusted
   for block width. A component from a wide block can be constructed to
   correlate with almost anything, particularly when the block has more
   features than children. Comparisons are valid *within* a block but not
   *between* blocks of very different widths. The proteome is notable because it
   achieves this correlation using only 36 named, interpretable proteins.

2. `plotVar` and `selectVar` identify selected features and show how each relates
   to the components. The circos plot also shows correlations **between features
   in different blocks**. Chords connect the features, and the cutoff limits the
   display to strong correlations. This reveals, for example, whether a selected
   protein varies with a selected urinary metabolite and whether both vary with
   the PCB congeners. The outer colours come from the `group` argument. This is
   a regression model with no groups; cohort was supplied only because
   `circosPlot()` requires a factor of the correct length. The colours are
   **not** part of the model.

3. The `keepX` values were fixed manually rather than evaluated
   by cross-validation. The conclusion is descriptive: *in these 896 children,
   the four molecular blocks share a direction of variation with organochlorine
   body burden, most visibly in the proteome and the serum metabolome, and the
   selected features carry
   that variation.* The model does **not** establish that exposure can be
   predicted in a new child. The two-block PLS1 model in Section 2 provides weak
   evidence for prediction, with Q2 =
   `r round(perf.pls.prot$measures$Q2$summary$mean[1], 3)` on one component.
:::

<!-- end solutions -->

## To go further

Some analysis extensions you can explore in your own time:

**Section 1, explore more blocks and characteristics.** Run `pca()` on each of
the other three omics blocks and the other eight exposure families (see
`names(helix$exposure)` that lists them). Recolour the sample plots by asthma outcome
(`helix$outcomes$asthma`) or maternal
education (`helix$outcomes$mat_edu`). Use `plot()` and `cor()` to compare a
component with the continuous covariates in `helix$covariates`. Which exposure
families measure one underlying pattern, as the organochlorines do, and which
contain several distinct patterns?

**Section 2, adjust for cohort.** Remove cohort from both the response and the
proteome using a simple linear model and extracting the residuals, then fit a new PLS model:

```{r 04-cohort-check, eval = FALSE}
Y.adj        <- residuals(lm(Y ~ helix$outcomes$cohort))
proteome.adj <- residuals(lm(proteome ~ helix$outcomes$cohort))

pls.adj <- pls(proteome.adj, Y.adj, ncomp = 3, mode = "regression")

plot(pls.adj$variates$X[, 1], Y.adj, pch = 16, col = "#009E7366",
     xlab = "proteome component 1, cohort removed",
     ylab = "summed PCBs, cohort removed")
abline(lm(Y.adj ~ pls.adj$variates$X[, 1]), lwd = 2)

cor(pls.adj$variates$X[, 1], Y.adj)
```

Does the association remain?

**Section 2, change the predictor or response.** Keep the same response and use
the serum, urine or transcriptome as the predictor block. Alternatively, keep
the proteome and change the response to `hs_dde_cadj_Log2`, a metal or a
phthalate. Which pairs produce a Q2 above the line?

**Section 3, change the response family.** Replace
`helix$exposure$organochlorines` in `X.list` with
`helix$exposure$air_pollution` or `helix$exposure$phthalates`. In the design
matrix code, change only the block name.

**Section 3, combine all exposure families.** Concatenate the nine families
into a single 81-variable response block:

```{r 04-pooled-env, eval = FALSE}
env <- do.call(cbind, helix$exposure)
dim(env)
```

Fit a sparse block PLS model that also selects on the Y block, then determine what the response variate represents.
Correlate
each of the selected response variables with the variate
and confirm whether those include the largest correlations.

**Section 3, vary the model.** Change the `keepX` values and compare the circos
plot and selected features. Then set every off-diagonal entry of the design
matrix to 1 and refit. What question does this model answer instead?

Further analyses and vignettes are available at
[mixOmics.org](https://mixOmics.org).

# Acknowledgement and required citation

The data used in this practical come from the HELIX study and were released by
ISGlobal for the Exposome Data Challenge 2021. Educational use is permitted.
The following citation paragraph must appear verbatim in any publication
arising from these data:

::: required-citation
This data were created as part of the ISGlobal Exposome data challenge 2021, presented in
this publication (preprint: https://arxiv.org/abs/2202.01680 - under review in Env. Int.).
The HELIX study [Vrijheid, Slama, et al. EHP 2014; Maitre et al. 2018 BMJ Open]
represents a collaborative project across six established and ongoing longitudinal
population-based birth cohort studies in six European countries (France, Greece,
Lithuania, Norway, Spain, and the United Kingdom). The research leading to these results
has received funding from the European Community's Seventh Framework Programme
(FP7/2007-2013) under grant agreement no 308333 – the HELIX project and the
H2020-EU.3.1.2. - Preventing Disease Programme under grant agreement no 874583 (ATHLETE
project). The data used for the analyses described in this manuscript were obtained from:
Figshare https://figshare.com/account/home#/projects/98813 (project number 98813 accesed
on MM/DD/YYYY) and github
https://github.com/isglobal-exposomeHub/ExposomeDataChallenge2021/.
:::

The extract distributed with this practical is shared for teaching with the
agreement of ISGlobal.

# References

::: {#refs}
:::

# Session Info

```{r }
sessionInfo()
```
